Universals of word order result from optimization of grammars for efficient communication

### 2019-03-24 second-draft/3-paper-names.tex
revising

% ,lin2017critical
\dropcap{U}nderstanding and explaining the universal properties of human languages is one of the central goals of Linguistics \cite{behaghel1909beziehungen,zipf1949human,greenberg1963universals,chomsky1993lectures,...}.
% \cite{behaghel1909beziehungen,zipf1949human,greenberg1963universals,lin2017critical} \cite{hawkins2007processing,bender2009linguistically,bender2013linguistic}
Across theoretical paradigms, linguists have hypothesized that language is shaped by efficiency in computation~\cite{berwick1984grammatical,chomsky2005three, hawkins2007processing} and communication~\cite{zipf1949human, Croft:Cruse:2004, Goldberg:2005}.
%Role of general computational constraints ~\cite{hauser2002faculty} % berwick2016only
In this work we examine this question from a computational perspective, demonstrating a fully formalized and implemented framework in which certain syntactic universals can be explained through the computational optimization of grammars for communicative efficiency. % information-theoretic

%An explanation for the universal properties of language would enable a deeper scientific understanding of what human language is and how to model it, with applications in psychology and natural language processing \cite{hawkins2007processing,bender2009linguistically,bender2013linguistic}. (TODO cite linguists)

@@

The contribution of our work is to show  that the word order of languages has evolved to optimize efficiency, and that efficiency optimization accounts for the Greenbergian word order correlations.
%In this work, we first provide a formalization of efficiency in the domain of word order. %, drawing on information-theoretic notions.

@@

In Information Theory, the usefulness of a communication channel depends on (1) how costly encoding and transmission is, and (2) how precisely messages can be recovered from codes (REF).

Eq.~\ref{eq:efficiency} is a special case of the objective function proposed in \cite{ferrericancho2002zipf,ferrericancho2007global,futrell2017memory} as a general objective for communicative systems, taking dependency structures as the underlying meanings to be conveyed.
Eq.~\ref{eq:efficiency} can also be seen as a simplified form of the Information Bottleneck \cite{tishby1999information}, a general objective function for lossy compression which has recently been applied to explain linguistic phenomena such as color naming systems \cite{zaslavsky2018efficient} (see SI section 4 for the precise relationship).

Efficiency is formalized as a weighted combination of terms representing the amount of information conveyed and the cost of communication.
%The \key{efficiency} of a language is a weighted combination
\begin{equation}\label{eq:efficiency}
	R_{\textit{Eff}} := R_{\textit{Pars}} + \lambda R_\textit{Pred}
\end{equation}
with an interpolation weight $\lambda \in [0,1)$. %In all experiments in this paper we use $\lambda=.9$ (see SI appendix section 5 for mathematical justification and for robustness to other choices).

In our model, these will depend on the grammar:
Grammars can differ in the degree to which a listener can unambiguously recover the dependency structure from an utterance.
This reflects the \key{parseability} of the language.
Maximizing the information provided by utterances about meanings is a standard formalization of Zipf's principle of listener economy~\cite{frank2012predicting,regier2015word}.
The degree to which listeners can reconstruct dependency structures from an utterance is formalized as the amount of information that utterances $u$ provide about their underlying dependency structures $t$:
\begin{equation}
	R_{Pars} := I[\mathcal{U},\mathcal{T}] = \sum_{t,u} p(t,u) \log \frac{p(t|u)}{p(t)}
\end{equation}
where the sum runs over all possible pairs of sentences $l$ and dependency structures $t$ in the language.
This quantity describes the degree to which dependency structures can be unambiguously recovered from sentences.
It is large when they can mostly be recovered, and small if sentences have high amounts of ambiguity about the dependency structure.
This quantity formalizes Zipf's Force of Diversification:
It can be maximized when trees can be decoded fully unambiguously from utterances.

@@

\begin{figure} % need Efficiency and (maybe) MLE langs on here
	\includegraphics[width=0.4\textwidth]{../results/correlations/figures/coverage-ground_eff-circles.pdf}
	\caption{per-languag predictions}\label{fig:per-lang}
\end{figure}

@@

%In order to test whether an objective function predicts a correlation, we selected all word order grammars created for the given function, and counted the percentage of grammars satisfying the correlation.
We conducted a multivariate mixed-effects logistic regression model predicting which correlations an optimized grammar satisfies, entering the language and language family of the underlying corpus as random effects.\footnote{We coded language families according to \url{universaldependencies.org}.}
%We are interested in the direction and significance of this effect: 
%If the effect is significant, in the positive direction, we can conclude that a correlation is predicted across corpora from languages belonging to different language families.
We estimated the posterior distribution of the number of correlations that the optimization process correctly predicts.
The estimated posterior probability that efficiency optimization predicts less than all eight correlations is $0.0242$. The probability that it predicts less than seven of the correlations is $<10^{-4}$.
%Predictability and parseability predict four and six correlations, respectively, making largely complementary predictions.

>>>>

% ,lin2017critical
\dropcap{U}nderstanding and explaining the universal properties of human languages is one of the central goals of Linguistics \cite{zipf1949human,greenberg1963universals,chomsky1993lectures,...}.
% behaghel1909beziehungen,
% \cite{behaghel1909beziehungen,zipf1949human,greenberg1963universals,lin2017critical} \cite{hawkins2007processing,bender2009linguistically,bender2013linguistic}
An explanation for the universal properties of language would enable a deeper scientific understanding of what human language is and how to model it, with applications in psychology and natural language processing \cite{hawkins2007processing,bender2009linguistically,bender2013linguistic}.
Across theoretical paradigms, linguists have hypothesized that language is shaped by efficiency in computation~\cite{berwick1984grammatical,chomsky2005three, hawkins2007processing} and communication~\cite{zipf1949human, Croft:Cruse:2004, Goldberg:2005}.
%Role of general computational constraints ~\cite{hauser2002faculty} % berwick2016only
In this work we examine this question from a computational perspective, demonstrating a fully formalized and implemented framework in which certain syntactic universals can be explained through the computational optimization of grammars for efficiency. % information-theoretic

@@

Drawing on data from 51 languages, we show  that the word order of languages has evolved to optimize efficiency, and that efficiency optimization accounts for the Greenbergian word order correlations.
%In this work, we first provide a formalization of efficiency in the domain of word order. %, drawing on information-theoretic notions.

@@

In Information Theory, the usefulness of a communication channel depends on (1) how costly encoding and transmission are, and (2) how precisely messages can be recovered from codes~\cite{shannon1948mathematical}.
This idea formalizes Zipf's forces of Unification and Diversification, and overall efficiency can be formalized as a weighted combination of terms representing the amount of information conveyed and the cost of communication:
%The \key{efficiency} of a language is a weighted combination
\begin{equation}\label{eq:efficiency}
	R_{\textit{Eff}} := R_{\textit{Informativity}} - \lambda R_\textit{Cost}
\end{equation}
with an interpolation weight $\lambda \in [0,1)$. %In all experiments in this paper we use $\lambda=.9$ (see SI appendix section 5 for mathematical justification and for robustness to other choices).
Such notions of efficiency have been proposed in \cite{ferrericancho2002zipf,ferrericancho2007global,futrell2017memory} as a general objective for communicative systems. %, taking dependency structures as the underlying meanings to be conveyed.
Eq.~\ref{eq:efficiency} can also be seen as a simplified form of the Information Bottleneck \cite{tishby1999information}, a general objective function for lossy compression which has recently been applied to explain linguistic phenomena such as color naming systems \cite{zaslavsky2018efficient} (see SI section 4 for the precise relationship). 
%Word meaning \cite{regier2015word}

In our model, both aspects of efficiency will depend on the grammar:
Grammars can differ in the degree to which a listener can unambiguously recover the dependency structure from an utterance.
This reflects the \key{parseability} of the language.
Maximizing the information provided by utterances about meanings is a standard formalization of Zipf's Force of Diversification~\cite{frank2012predicting,regier2015word}. % also goodman2013knowledge,
The degree to which listeners can reconstruct dependency structures from an utterance is formalized as the amount of information that utterances $u$ provide about their underlying dependency structures $t$:
\begin{equation}
	R_{Pars} := I[\mathcal{U},\mathcal{T}] = \sum_{t,u} p(t,u) \log \frac{p(t|u)}{p(t)}
\end{equation}
where the sum runs over all possible pairs of sentences $l$ and dependency structures $t$ in the language.
This quantity describes the degree to which dependency structures can be unambiguously recovered from sentences.
It is large when they can mostly be recovered, and small if sentences have high amounts of ambiguity about the dependency structure.
In line with Zipf's Force of Diversification, this quantity is maximized when trees can be decoded fully unambiguously from utterances.

@@

\begin{figure} % need Efficiency and (maybe) MLE langs on here
	\includegraphics[width=0.3\textwidth]{../results/correlations/figures/coverage-ground_eff-circles.pdf}
	\caption{Across 51 languages, efficiency optimization predicts Greenberg's correlations. For every one of the 51 languages, we show which universals are satisfied in the actual grammar (left), and in the grammar optimized for efficiency (right). Efficiency optimization replicates the universals at rates similar to the real grammars.}\label{fig:per-lang}
\end{figure}

@@

%In order to test whether an objective function predicts a correlation, we selected all word order grammars created for the given function, and counted the percentage of grammars satisfying the correlation.
We conducted a multivariate mixed-effects logistic regression model predicting which correlations an optimized grammar satisfies, entering the language and language family of the underlying corpus as random effects.\footnote{We coded language families according to \url{universaldependencies.org}.}
%We are interested in the direction and significance of this effect: 
%If the effect is significant, in the positive direction, we can conclude that a correlation is predicted across corpora from languages belonging to different language families.
We estimated the posterior distribution of the number of correlations that the optimization process correctly predicts.
The estimated posterior probability that efficiency optimization predicts less than all eight correlations is $0.0242$. The probability that it predicts less than seven of the correlations is $<10^{-4}$.

We conducted the same analysis for grammars optimized for predictability and parseability separately.
Predictability and parseability predict four and six correlations, respectively, making largely complementary predictions.
